EBASE manuscript

### 2023-01-11 manu-draft.qmd
optex results text added

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each.  As for the BASEmetab model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can of course be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.  The default priors were based on those that returned reasonable parameter estimates using a forward-feeding oxygen mass balance equation call FWOXY [ForWard OXYgen, @Arriola22].  For FWOXY, a dissolved oxygen time series is simulated by providing known parameter values in an equation similar to @eq-ebase. All of the physical parameters are held constant, e.g., wind speed, temperature, and salinity.  The other required inputs are the $a$, $r$, and $b$ parameters, which are set by the user.  An initial oxygen concentration is provided and the simulated time series is created by forward addition of the inputs. As such, the dissolved oxygen output simulated by FWOXY as change per unit time were modelled by EBASE using the constant physical parameters as input. Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters from FWOXY:

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a simulated time series created with FWOXY. These assessments evaluated if 1) EBASE estimates similar values for the known parameters, and 2) how the estimated values change with different model configurations. The simulated time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida.  The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), and salinity (psu).  Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation). An oxygen time series was generated from Fwoxy using the inputs from Apalachicola Bay ([@fig-simapa]a), such that the simulated dissolved oxygen was created as close as possible to the observed given the inputs.  Known values for $a$ and $r$ parameters were returned for each time step of the simulated data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-simapa]b and c).  The only parameter that could not be simulated by Fwoxy was $b$, which was set to 0.251 $\buni$.

@@

Several conclusions can be made from the results in @fig-apasumdat based on the patterns of the $R^2$ values. Overall, the model was able to estimate the $a$ parameter with reasonable precision regardless of any of the prior distributions, whereas most of the variation was observed in the $r$ (respiration, R) and $b$ parameters.  Changing the variance of the $a$ parameter had little effect on any of the comparisons.  Smaller variance of the prior distribution for the $r$ parameter generally contributed to lower $R^2$ for all parameters, with the precision being worse for the 7-day optimization period (e.g., the $a$ parameter).  Larger variance of the prior distribution for the $b$ parameter generally contributed to lower $R^2$ only for the gas exchange estimate (D), although this effect was minimized with the 7-day optimization period.  As such, the results suggest that the default prior distributions used in EBASE may be appropriate in some cases, but increasing the prior variance for $a$ and $r$ and decreasing the variance for $b$ will likely produce more optimal solutions. The actual model results for the best performing combination of prior distributions for the one-day and 7-day optimization is shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability).

@@

![Comparison of EBASE results with known metabolic parameters from a simulated time series. Results are compared for unique combination of prior distributions, where the variances are evaluated from low (L), medium (M), to high (H) for a one-day and seven-day model optimization period. The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and simulated values.  The $b$ parameter is not evaluated because it is a fixed constant for the simulated data.](figs/apasumdat.png){#fig-apasumdat}

>>>>

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each.  As for the BASEmetab model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can of course be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.  The default priors were based on those that returned reasonable parameter estimates using a forward-feeding oxygen mass balance equation called Fwoxy [ForWard OXYgen, @Arriola22].  For Fwoxy, a dissolved oxygen time series is simulated by providing known parameter values in an equation similar to @eq-ebase. All of the physical parameters are held constant, e.g., wind speed, temperature, and salinity.  The other required inputs are the $a$, $r$, and $b$ parameters, which are set by the user.  An initial oxygen concentration is provided and the simulated time series is created by forward addition of the inputs. As such, the dissolved oxygen output simulated by Fwoxy as change per unit time were modelled by EBASE using the constant physical parameters as input. Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters from Fwoxy:

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a simulated time series created with Fwoxy. These assessments evaluated if 1) EBASE estimates similar values for the known parameters, and 2) how the estimated values change with different model configurations. The simulated time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida.  The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), and salinity (psu).  Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation). An oxygen time series was generated from Fwoxy using the inputs from Apalachicola Bay ([@fig-simapa]a), such that the simulated dissolved oxygen was created as close as possible to the observed given the inputs.  Known values for $a$ and $r$ parameters were returned for each time step of the simulated data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-simapa]b and c).  The only parameter that could not be simulated by Fwoxy was $b$, which was set to 0.251 $\buni$.

@@

Several conclusions can be made from the results in @fig-apasumdat based on the patterns of the $R^2$ values. Overall, the model was able to estimate the $a$ parameter with reasonable precision regardless of any of the prior distributions, whereas most of the variation was observed in the $r$ (respiration, R) and $b$ parameters.  Changing the variance of the $a$ parameter had little effect on any of the comparisons.  Smaller variance of the prior distribution for the $r$ parameter generally contributed to lower $R^2$ for all parameters, with the precision being worse for the 7-day optimization period (e.g., the $a$ parameter).  Larger variance of the prior distribution for the $b$ parameter generally contributed to lower $R^2$ only for the gas exchange estimate (D), although this effect was minimized with the 7-day optimization period.  As such, the results suggest that the default prior distributions used in EBASE may be appropriate in some cases, but increasing the prior variance for $a$ and $r$ and decreasing the variance for $b$ will likely produce more optimal solutions. The actual model results for the best performing combination of prior distributions for the one-day and 7-day optimization are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability).  The best performing models were those with the highest mean $R^2$ values across all parameter estimates in @fig-apasumdat.  For both models, the metabolic parameters were estimated with high precision, and in some cases nearly perfect precision (i.e., dissolved oxygen and gas exchange, D).  Estimates for the fixed $b$ parameter were most variable, but not biased as the mean for both models was close to 0.252 $\buni$.

@@

![Comparison of EBASE results with known metabolic results from a simulated time series created with Fwoxy. Results are compared for unique combination of prior distributions, where the variances are evaluated from low (L), medium (M), to high (H) for a one-day and seven-day model optimization period. The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and simulated values.  The $b$ parameter is not evaluated because it is a fixed constant for the simulated data.](figs/apasumdat.png){#fig-apasumdat}

![Comparison of EBASE results for the best performing model with known metabolic results from a simulated time series created with Fwoxy. The best performing model for the one-day (left, ndays = 1) and seven-day (right, ndays = 7) optimization period is based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-apasumdat).  The standard deviations of the prior distributions of each parameter for the best performing model are shown at the top.  The results are averaged at the time step of the model optimization period.  The $b$ parameter is fixed at 0.251 $\buni$.](figs/optex.png){#fig-optex fig-align="center"}
